Lending Club loans 
========================
2,500 peer-to-peer loans issued through the Lending Club 
(https://www.lendingclub.com/home.action). The interest rate of these loans is determined 
by the Lending Club on the basis of characteristics of the person asking for the loan such as 
their employment history, credit history, and creditworthiness scores. 
The purpose of your analysis is to identify and quantify associations between the interest rate
of the loan and the other variables in the data set. In particular, you should consider whether
any of these variables have an important association with interest rate after taking into 
account the applicant s FICO score. For example, if two people have the same FICO score, 
can the other variables explain a difference in interest rate between them?
A FICO score is a US credit rating, a measure of creditworthiness. A higher score indicates that FICO, a private credit rating agency, thinks the potential borrower is more likely to pay back their loans.
--------
```{r}
library(Hmisc)
library(stringr)
setwd("G:/PhDcourses/dataanalysis/works/firstassignment/data")
a<-read.csv("./loansData2.csv")
lmstep<-lm(a$Interest.Rate~.,data=a)
ss<-step(lmstep)
print(date())
ss
#type, head and name of variables
str(a)
#min, max, median and quantile of variables
summary(a)
#finding missing values
colSums(is.na(a))
```
**Exploratory analysis**
```{r}
hist(a$Interest.Rate,xlab="Interest.Rate")

x<-data.frame(str_split_fixed(a$FICO.Range,"-",2))
x$X1 <- as.numeric(as.character(x$X1))
x$X2 <- as.numeric(as.character(x$X2))
ficoMean<-rowMeans(x)
a<- data.frame(a,ficoMean)
anova(lm(a$Interest.Rate~a$ficoMean))
data2<-cbind(a$Amount.Requested,a$Amount.Funded.By.Investors,a$Interest.Rate,a$Loan.Length,a$Loan.Purpose,a$Debt.To.Income.Ratio,a$State,a$Home.Ownership,a$ Monthly.Income,a$Open.CREDIT.Lines,a$Revolving.CREDIT.Balance,a$Inquiries.in.the.Last.6.,a$Employment.Length )
#you should only use PCA when your variables were not highly correlated http://www.youtube.com/watch?v=Heh7Nv4qimU
pairs(data2)
requestedQuartile <- as.integer(cut(a$Amount.Requested , quantile(a$Amount.Requested , probs=0:4/4), include.lowest=TRUE))
plot(a$Interest.Rate, a$ficoMean, col=(a$Loan.Length==36)+1, pch=19, xlab="Interest Rate (%)", ylab="FICO score")
legend(x="topright", c("36 months", "60 months"), col=c("red", "black"), pch=c(19,19))
####################
plot(a$Interest.Rate, a$ficoMean, col=requestedQuartile, pch=19, xlab="Interest Rate (%)", ylab="FICO score")
legend(x="topright", c("Amount Requested 1st quartile", "Amount Requested 2nd quartile", "Amount Requested 3rd quartile", "Amount Requested 4th quartile"), col=seq(1:4), pch=c(19,19))
#######################
table(a$Loan.Length)
anova(lm(a$Interest.Rate  ~.,a))

ml1<-lm(a$Interest.Rate~a$low)
confint(ml1,level=0.95)
plot(a$low,a$Interest.Rate,pch=19,col=as.factor(a$Loan.Length),xlab="FICO Value",ylab="Interest rate", main="fitted linear model overlayed with Interest rate based on FICO value colored by loan length")
points(a$low,ml1$fitted,col="green",pch="-",cex=5)
summary(ml1)
par(mfrow = c(1, 2))
plot(a$low,ml1$residuals,pch=19,col=as.factor(a$Loan.Length),xlab="FICO Value",ylab="model residual")
plot(ml1$fitted,ml1$residuals,pch=19,col=as.factor(a$Loan.Length),xlab="fitted values",ylab="model residual")
#color by Debt.To.Income.Ratio
incut = cut2(a$Debt.To.Income.Ratio, g = 5)
par(mfrow = c(1, 2))
plot(a$low,ml1$residuals,pch=19,col=incut)
plot(ml1$fitted,ml1$residuals,pch=19,col=incut)

#this shows that the higher FICO range the lower interst rate. the interst rate distribution in higher FICO range has smaller quantile interval meaning lower variance and more skewd. the interst rate distribution in lower FICO range has larger quantile interval meaning higher variance. the number of people with high FICO is also wery low. also the loan.length is a good seprator so it worth adding it to model.

ml2<-lm(a$Interest.Rate~a$low+as.factor(a$Loan.Length))
summary(ml2)
confint(ml2,level=0.95)
par(mfrow = c(1, 2))
plot(a$low,ml2$residuals,pch=19,col=as.factor(a$Loan.Length))
plot(ml2$fitted,ml1$residuals,pch=19,col=as.factor(a$Loan.Length))
############color by Debt.To.Income.Ratio
incut = cut2(a$Debt.To.Income.Ratio, g = 5)
par(mfrow = c(1, 2))
plot(a$low,ml2$residuals,pch=19,col=incut)
plot(ml2$fitted,ml1$residuals,pch=19,col=incut)
############

plot(a$Interest.Rate~jitter(as.numeric(a$Loan.Length)),col="green",xaxt="n")
axis(side=1,at=unique(as.numeric(as.factor(a$Loan.Length))),lables=unique(as.factor(a$Loan.Length)))
meanIR<-tapply(a$Interest.Rate,a$Loan.Length,mean)
str(meanIR)
points(1:length(meanIR),meanIR,col="red",pch="-",cex=5)

hist(a$Amount.Requested,breaks=100)
quantile(a$Amount.Requested)
boxplot(a$Amount.Requested~a$Home.Ownership,varwidth=TRUE)
#Monthly.Income
hist(a$Monthly.Income,breaks=10)
quantile(a$Monthly.Income)
boxplot(a$Monthly.Income)
#Open.CREDIT.Lines
hist(a$Open.CREDIT.Lines)
quantile(a$Open.CREDIT.Lines)
boxplot(a$Open.CREDIT.Lines)
#Revolving.CREDIT.Balance
hist(a$Revolving.CREDIT.Balance)
quantile(a$Revolving.CREDIT.Balance)
boxplot(a$Revolving.CREDIT.Balance)
#Inquiries.in.the.Last.6.Months
hist(a$Inquiries.in.the.Last.6.Months,breaks=100)
quantile(a$Inquiries.in.the.Last.6.Months)
boxplot(a$Inquiries.in.the.Last.6.Months)

```
**univariate tables **
```{r}
table(a$Loan.Purpose)
table(a$State)
table(a$Home.Ownership)
table(a$Loan.Length)
hist(a$Debt.To.Income.Ratio)
hist(a$Interest.Rate)
table(a$FICO.Range)
table(a$Employment.Length)
```
**correlations plot**
```{r}
boxplot(a$Interest.Rate~a$FICO.Range,varwidth=TRUE)
plot(a$Interest.Rate,a$Debt.To.Income.Ratio,col=a$Home.Ownership)
plot(a$Interest.Rate,a$Debt.To.Income.Ratio,col=a$FICO.Range)
#plot(a$Interest.Rate,a$)
```
**principal component analysis**
```{r}
#first spcify the variable you want to predict and if it is not a factor, cut it into intervals coerce factros
numericIR<-as.numeric(as.factor(cut2(a$Interest.Rate,g=5)))
#since FICO range "660-664" has 125 individual, we diecided to do PCA on that and choose the best column who seprated columns well but it seems that none of the columns is a good seprator ....
plot(a[a$FICO.Range=="660-664",6],pch=19,cex=5,col=numericIR)
#here we have all the entries in the x axis and if we can find a column that at least seprate colors into 2 or more groups horizontally then that variable is usefull to classify those color groups. but unfortunatly no one is useful here. so we may try SVD
colSums(is.na(a))
which(is.na(a[,9]))
which(is.na(a[,11]))
which(is.na(a[,12]))
which(is.na(a[,13]))

b<-a[-c(367,1595),]
colSums(is.na(b))
svd1<-svd(scale(b[b$FICO.Range=="665-669",-c(3,5,7,8,10,14)]))

#svd1=svd(scale(a[,-c(3,5,7,8,10,14)]))
plot(svd1$u[,1],col=numericIR,pch=19)
plot(svd1$u[,3],col=numericIR,pch=19)
#according to the plots svds also are unable to diffrentiate among intervals for IR :-(

lm1<-lm(a$Interest.Rate~a$Amount.Requested)
lm1
plot(a$Amount.Requested,a$Interest.Rate,col=as.numeric(a$FICO.Range),pch=19)
points(a$Amount.Requested,lm1$fitted,col="red")
plot(a$Amount.Requested,lm1$residuals,col=as.numeric(a$FICO.Range),pch=19)
plot(lm1$fitted,lm1$residuals,col=as.numeric(a$FICO.Range),pch=19)

###########Monthly.Income
milog<-log(a[!is.na(a$Monthly.Income),c("Monthly.Income")]+1)
lm2<-lm(a$Interest.Rate~a$Monthly.Income)
lm2
plot(log(a$Monthly.Income),a$Interest.Rate)#,col=as.numeric(a$FICO.Range),pch=19)
points(milog,lm2$fitted,col="red")
plot(milog,lm2$residuals,col=as.numeric(a$FICO.Range),pch=19)
plot(log(lm2$fitted+1),lm2$residuals,col=as.numeric(a$FICO.Range),pch=19)
##############
###########Loan.Length
ccc<-a[a$FICO.Range=="665-669",]
boxplot(ccc$Interest.Rate~ccc$Loan.Length,xlab="loan length in month",ylab="intertest rate")
boxplot(a$Interest.Rate~a$Loan.Purpose,xlab="loan purpose",ylab="intertest rate")
#this shows that loan length is a very good indiicator for interst rate.
lm3<-lm(a$Interest.Rate~as.numeric(as.factor(a$FICO.Range)))
lm3
plot(as.numeric(as.factor(a$FICO.Range)),a$Interest.Rate,col=as.factor(a$Loan.Length),pch=19)
points(as.numeric(as.factor(a$FICO.Range)),lm3$fitted,col="red")
plot(as.numeric(as.factor(a$FICO.Range)),lm3$residuals,col=as.factor(a$Loan.Length),pch=19)
plot(lm3$fitted,lm1$residuals,col=as.factor(a$Loan.Length),pch=19)
#plots are representative that loan length is totally 
lmf<-lm(a$Interest.Rate~a$low)
summary(lmf)
confint(lmf)
plot(a$FICO, lmf$residuals, pch = 19, col=as.factor(a$Loan.Purpose))
plot(lmf$fitted, lmf$residuals, pch = 19, col = as.factor(a$Loan.Purpose))

######
lmm<-lm(a$Interest.Rate~as.factor(a$Loan.Length))
summary(lmm)
confint(lmm)
plot(a$FICO, lmf$residuals, pch = 19, col=as.factor(a$Loan.Purpose))
plot(lmf$fitted, lmf$residuals, pch = 19, col = as.factor(a$Loan.Purpose))
##################
plot(a$Interest.Rate,a$Open.CREDIT.Lines)
plot(a$Interest.Rate,a$Revolving.CREDIT.Balance)
plot(a$Interest.Rate,a$Inquiries.in.the.Last.6.Months)
#plot(a$Interest.Rate,a$)
#plot(a$Interest.Rate,a$)
```
**regression with factor variable**
```{r}
#modeling IR based on FICO
factorlm<-lm(a$Interest.Rate~as.factor(a$FICO.Range))
confint(factorlm)
#as we can see the confidance interval for difference among "640-644" FICO and first 12 FICO rating covers 0 so there is no big difference among IR of "640-644" and first other 12 FICO rating.
summary(factorlm)
#computing anova will help us to understand if IR is associated with FICO at any level.
#anova is the most common approach to determin how factor variable is assiciated with the quantitative variables
anova(factorlm)
#the realtion among FICO-range and Intesrt rate is statistically significant according to p value reported and high f-value(meaning that )
#to plot all possible differences between pairs of factor levels we use Tukey
lmtuk<-aov(a$Interest.Rate~as.factor(a$FICO.Range))
TukeyHSD(lmtuk)
################
#modeling IR based on home ownership
flm<-lm(a$Interest.Rate~as.factor(a$Home.Ownership))
confint(flm)
#as we can see the confidance interval for IR difference among "MORTGAGE" and all others except "RENT" covers 0 so there is no big difference among IR of "MORTGAGE" and others except "RENT"
summary(flm)
#computing anova will help us to understand if IR is associated with homeownership at any level.
#anova is the most common approach to determin how factor variable is assiciated with the quantitative variables
anova(flm)
#the realtion among home-ownership and IR is statistically significant according to p value reported but f-value is not that high
#to plot all possible differences between pairs of ownership levels we use Tukey
lmt<-aov(a$Interest.Rate~as.factor(a$Home.Ownership))
TukeyHSD(lmt)
##########
#**Loan.Purpose**
#modeling IR based on Loan.Purpose
flmp<-lm(a$Interest.Rate~as.factor(a$Loan.Purpose))
confint(flmp)
#those confidance intervals who covers 0 is not that different from car loan
summary(flmp)
#computing anova will help us to understand if IR is associated with loan purpose at any level.
#anova is the most common approach to determin how factor variable is assiciated with the quantitative variables
anova(flmp)
#the realtion among loan purpose and IR is statistically significant according to p value reported but f-value is not that high
#to plot all possible differences between pairs of loans' purposes levels we use Tukey
lmt<-aov(a$Interest.Rate~as.factor(a$Loan.Purpose))
TukeyHSD(lmt)
##########
```
**multiple regression with factor variable**
```{r}
plot(a$Interest.Rate,a$Monthly.Income,col=as.factor(a$FICO.Range),pch=19)
#looks weired , lets try transform by log :)
logIncome<-log(a$Monthly.Income)
plot(a$Interest.Rate,logIncome,col=as.factor(a$FICO.Range),pch=19)
#it doesn't seem there is a direct relation among those variables
lm<-lm(a$Interest.Rate~a$Monthly.Income)
summary(lm)
#our factor variable have multiple values so


```
**modeling FICO value**
```{r}
plot(a$low,log(a$Monthly.Income),pch=19,col="blue")
plot(a$low,a$Open.CREDIT.Lines,pch=19,col="blue")
plot(a$low,log(a$Revolving.CREDIT.Balance),pch=19,col="blue")
plot(a$low,a$Inquiries.in.the.Last.6.,pch=19,col="blue")
boxplot(a$low~a$Employment.Length)
boxplot(a$low~a$State)
boxplot(a$low,a$Home.Ownership)
hist(a$low,xlab="FICO value",main="histogram of FICO value")
     str(a[a$low<700,])
str(a[a$low>725,])
quantile(a$low)

```
